Insight Into Lumbar Back Pain \subtitle{What the Lumbar Spine Tells About Your Life}

### 2014-08-29 draft.tex
Revvised Paper

\section{\uppercase{Epidemiological Background}}
\label{sec:EpidemiologicalBackground}
%\noindent \com{Background. Epidemiological Workflow, focus on statistical resilience, image data hard to analyze due to the large amount of subjects, poor image quality and lack of methods}.
\noindent Epidemiology is the study of dissemination, causes and results of health-related states and events.
%%
Epidemiological reasoning relies on a strict statistically driven workflow \cite{Fletcher}:
%%
\begin{itemize}
%%
	\item Physicians formulate hypotheses based on observations made in their clinical practice.
%%
	\item To assess a hypothesis, epidemiologists compile a list of variables depicting it.
%%
	\item Statistical methods, such as regression analysis, assess the association of selected variables with the investigated disease.
%%
\end{itemize}
%%
Mutually dependent variables make this analysis challenging.
%%
Many diseases, such as different cancer types, are more likely with increasing age.
%%
When for example analyzing influences of nutrition to prostate cancer, the results need to be age-normalized.
%%
Age acts as an \emph{confounder} for prostate cancer.
%%
These \emph{confounding} variables are often hard to find.
%%
Statistical correlation does not imply causation--epidemiologists need to assess the medical soundness of the statistical results.
%%------------------------------
\subsection{Epidemiological Data}
Epidemiological data can stem from a wide range of studies.
%%
The study type depends on the condition of interest.
%%
Most common are case-control studies, analyzing one specific disease and its influences.
%%
We focus on data from large scale cohort studies.
%%
These studies aim to collect as much data as possible for each subject.
%%
As a result, these data can be analyzed regarding many diseases and conditions.
%%

Epidemiological data is heterogenous and incomplete.
%%
Women-specific question can only be acquired for female subjects, data about a disease treatment only for subjects suffering from this condition.
%%
Therefore, statistical analysis has to take missing data into account.
%%

@@

Modern cohort studies often comprise medical image data.
%%
These data are hard to analyze, since segmentation algorithms are not generally available and need to be custom-made for each body structure.
%%
Segmentation data is usually analyzed by abstracting it into key figures, such as diameters or distances.
%%
These numeric values can be compared with non-image variables to retrieve correlations.
%%--------------------------------
\subsection{Lower Back Pain}
%%
The lower (\emph{lumbar}) spine is the most stressed part of the spine.
%%
Lower back pain is one of the most frequent diseases in the western civilization.
%%
Epidemiologists assume associations between lumbar back pain and lifestyle factors.
%%
These include nutrition, sporting activities and body posture at work.
%%
The exact causes as well as particularly vulnerable risk groups are not known.
%%
Potential \emph{confounding} effects are also subject of current research.
%%

Epidemiologists want to characterize the healthy aging process of the spine.
%%
To achieve this, they have to analyze the lumbar spine shape as well as the mentioned lifestyle factors.
%% -------------------------------
\section{\uppercase{Related Work}}
\label{sec:RelatedWork}
%%
\paragraph{Visual Analysis of Image and non-Image Data}
%%
Klemm et al. \cite{Klemm2014VIS} propose an extension to the existing epidemiological workflow using Interactive Visual Analysis (IVA).
%%
The workflow consists of an iterative sequence of group selection, variable selection and visualization.
%%
It aims to trigger \emph{hypothesis generation} by providing visualizations able to concurrently analyzing multiple heterogenous variables at once using correlation measures.
%%
We incorporate the IVA approach by concurrently displaying complex relationships between variables using decision trees.
%%
Steenwijk et al. \cite{Steenwijk} propose a coordinated linked view system for both image-related and non-image data.
%%
It incorporates multiple plot types, such as scatter plots, parallel coordinates and time plots with brushing and linking facilities to successively analyze a data set.
%%
Similar to our work, they quantify image data to map them into the same information space as non-image data.
%%
Turkay et al. \cite{Turkay} follow a similar approach by deriving descriptive data metrics from image data.
%%
Their proposed \emph{deviation plot} shows distribution-specific measures of a variable, such as skewness or inter-quantile-range, making dimensions comparable.
%%
This approach aims to trigger \emph{hypothesis generation} by outlining tendencies between these dimensions.

\paragraph{Decision Rule Driven Analysis of Medical Data.}
Closest to our work is the work of Gla{\ss}er \cite{Glasser2013} and Niemann et al. \cite{Niemann2014}.
%%
Related work in this field often focuses on clinical diagnosis.
%%
Gla{\ss}er et al. \cite{Glasser2013} use parameters derived from DCE-MRI image data to classify breast tumors.
%%
Their parameters are derived from clustering of bloodflow in the tumorous tissue.
%%
These features trains a decision-tree classifier, concluding that the extracted features alone are not sufficient for tumor type classification.
%%
Niemann et al. \cite{Niemann2014} assess hepatic steatosis (fatty liver disease) risk factors using decision trees.
%%
They presented an interactive data mining tool, which can analyze association rules and highlight interesting relations, which may trigger new hypothesis about the data.
%%
The presented paper either validate the significance of image derived parameter or derive new relationships by training decision trees to explain target variables.
%%

Pinhero et al. \cite{Pinheiro} applied association rule mining to create high-confidence decision rules about the mortality rate in liver cancer data.
%%
Decision rules are easily readable and understandable and are therefore well suited for classifying medical data.
%%
The results can be easily communicated to domain experts with little knowledge about the algorithm behind the results.
%%
Once these rules, however, get too complex, analyzing these decision rules can get complex.
%%
Sekhavat and Hoeber \cite{Sekhavat} propose a interactive visualization techniques for decision trees, where all decisions are displayed in a heat map
%%
The visualization translates the decision tree into a heat map, where each cell represents the connectivity between a initial value and an target value.
%%
The resulting interactive visualization is well suited for analyzing a decision tree, but not for multiple trees.
%%
\paragraph{Visual Analysis of Heterogenous Non-Image Data.}
%%
Zhang et al. \cite{Zhang} analyze subject groups in a web-based linked view system.
%%
The resulting decision rules aim to categorize new subjects as they are added to the data.
%%
They define a cohort as parameter-divided subject group, differing from the epidemiological understanding of a cohort.
%%
The paper lacks details about handling of missing data, the definition of similarity or which statistical measures are used.
%%
Generalized Pairs Plots (\texttt{GPLOM'S}) visualize heterogenous variables in a plot matrix \cite{GPLOMS, Francois2013}.
%%
The plot depends on the type of variables, which are pairwise visualized.
%%
The plot matrix is well suited for an overview visualization, but takes up much screen space and is therefore only suitable for few variables at once (see Fig.~\ref{image-parameter-range}).
%%
Dai et al. \cite{Dai} defined a \emph{Concept Map} linking cancer-associated features by incorporating choropleth maps with bar charts, parallel coordinates and scatter plots with regression lines.
%%
The \emph{Concept Map} is iteratively refined using primarily the underlying geographical data.
%%
Our work does not contain geographical information of any kind.
%%----------------------------
\paragraph{SHIP Data Analysis.}
Klemm et al. analyzed the lumbar spine variability of of 490 \texttt{SHIP-2} subjects \cite{Klemm2013VMV}.
%%
They incorporate hierarchical agglomerative clustering to derive shape groups, yielding several groups of average shape and several outlier cluster.
%%
The extracted lumbar spine shape was correlated with subject size.
%%
Out data extraction process is based on the detection algorithm and lumbar spine canal extraction presented in this paper.
%%
Clustering techniques applied to non-image data are strongly dependent on the chosen variables and distance measures.
%%
Klemm et al. found \texttt{k-Prototypes} and \texttt{DBSCAN} appropriate for this task \cite{Klemm2014BVM}.
%%
\\
Unique for our work is the combination of an IVA-approach by observing interesting feature relations in context of image-derived features using multiple decision trees for one visualization.
%%
We abstract the decision tree results similar to \cite{Turkay} to make them comparable in a overview visualization.
%%
\section{The Lumbar Spine Data Set}
\label{sec:MaterialsAndMethod}
%%
\noindent Our approach allows to analyze many variables simultaneously.
%%
Therefore, epidemiologists compiled the data set with a with a wide range of variables possibly correlating with lumbar back pain.
%%
The data set comprises of 6,753 subjects from two cohorts (4,420 from \texttt{SHIP-Trend-0} and 2,333 from \texttt{SHIP-2}).
%%
%It is an active research area to determine these correlations.
%%

%%
\subsection{Non-Image Data}
%\com{Analysis Using Heterogeneous Correlation}
The variables range from somatometric variables describing body measures to medical examinations, such as laboratory tests as well as lifestyle factors as sport activity or nutrition.

@@

A hierarchical finite element method was used to detect the lumbar spine in the MRI data \cite{Rak2013}.
%%
The tetrahedron-based finite element model is initialized using three user-placed land marks.
%%
A click on the L3 vertebra center gives the model the initial position, two clicks on the top and bottom of the vertebra describes an initial model height estimation.
%%
The registered model captures information about the lumbar spine canal shape and the position of the L1-L5 vertebrae \cite{Klemm2013VMV} (Fig.~\ref{fig:centerline}~a).
%%
The detection fails for several subjects due imprecise initialization, imaging artifacts or strongly deformed spines.
%%
$2,540$ lumbar spine models were obtained and constitute the foundation of this work.
%%
%%
% \com{From VIS'14 Paper}
% The lumbar spine was detected in the image data using a hierarchical finite element method by Rak et al. [32].
% %%
% This semiautomatic method requires the user to initialize the tetrahedron-based finite element models (FEM) with a click on the L3 vertebra.
% %%
% Two user-defined landmarks on the top and bottom of the L3 vertebra describe an initial model height estimation.
% %%
% The model uses a weighted sum of T1 and T2-weighted MR images to detect the lumbar spine shape.
% %%
% Once registered, it captures information about the shape of the lumbar spine canal as well as the position of the L1-L5 vertebrae [21].
% %%
% Due to incorrect initialization, strongly deformed spines, contrast differences and artifacts, the model was not able to detect lumbar spines for all subjects.
% %%
% We obtained and worked with 2,540 tetrahedron models of the lumbar spine.
% %%
% For clustering, we extracted the centerline of the lumbar spine canal, which captures information about lordosis and scoliosis (the medical terms for spine curvature) [21].
% %%

We have to assess the model accuracy to extract key figures from it.
%%
The detection model depicts the vertebrae positions, but lacks detailed information about their volume.
%%
It captures reliable information about spine canal curvature.
%%
In a previous work \cite{Klemm2013VMV,Klemm2014VIS}, we extracted a centerline representation of the lumbar spine canal from the detection model (Fig.~\ref{fig:centerline}~a).
%%
Using the Frenet formulas \cite{Frenet}, we extracted the following metrics from the model (Fig.~\ref{fig:centerline}~b):
%%
%Thus, we extracted the following metrics from the model:
%%
%\paragraph{Curvature}
\begin{itemize}
	% \item The curvature of the centerline is calculated using the summed mean curvature between adjacent points:
% %%
% $\sum \frac{curvature_i}{curvature_all}$.
	\item \emph{Curvature} is calculated as weighted sum of curvature between all adjacent points describing the centerline: $\sum \frac{curvature_i}{curvature_{all}}$.
%%
	\item \emph{Torsion} (deviation of a curve from its current course) is calculated as weighted sum of torsion between all adjacent points describing the centerline: $\sum \frac{torsion_i}{torsion_{all}}$.
	\item \emph{Curvature angle $\alpha$} is defined by the middle point of the spine canal centerline as \emph{vertex} and the line between middle point and top/bottom point as \emph{sides}.
%%
\end{itemize}
%%
These figures are also extracted in the sagittal, coronal and transversal projection of the model.
%%
We assess the information gain of each dimension using these projections.
%%
This gives us a total of 9 image-derived parameters.

@@

Larger people have also a longer spine and its shape is more straight compared to smaller subjects.

@@

\noindent As first experiment we correlated the shape parameter with the binary back pain indicator using a generalized pairs plot \cite{GPLOMS}.

@@

Pairs plots visualize continuous and categorical variables pairwise in a matrix view.

@@

Scatterplots display pairs of continuous variables, bar charts for continuous variable as a function of a categorical variable and box plots pairs of categorical variables.

@@

Curvature, curvature angle and torsion correlate strongly with their planar projections.

@@

Also the mean curvature and the curvature angle correlate by a factor of -0.89.

@@

\subsubsection{Assessing the Information Gain Using the PCA}
%%
To determine the information gain per image-derived variables, we calculated a PCA and compared the loadings per dimension.
%%
The first three principal components explain 75\% of variance in the image-derived variables.
%%
The first principal component explains 47\% of the variance and weights primarily \emph{mean curvature}, \emph{curvature sagittal}, \emph{curvature angle} and \emph{curvature angle sagittal}.
%%
The second component, adding 16\% of variance, weigths  \emph{curvature coronal} and \emph{curvature angle coronal}.
%%
The third component explains 12\% of variance and weights \emph{torsion} and \emph{curvature transverse}.
%%
This supports our prior conclusion about the low information gain of the transverse planes.
%%
\emph{Torsion} also adds little variance to the information space.
%% ------------------------------------------
\subsubsection{Heterogenous Correlations}
%%
We then expanded our focus on correlations of image-derived parameters with all other non-image variables.
%%
We applied a heterogenous correlation technique to derive correlations between all variables is the data set.
%%
The method uses the following correlation metrics for the different type combination:
%%
\begin{itemize}
%%
\item \emph{Pearson product-moment} for two continuous variables, 
\item \emph{Polyserial} correlation for one continuous and one categorical variable, and
\item \emph{Polychoric} correlations for two categorical variables.
\end{itemize}
\com{TODO: Abbildung einer Korrelationsmatrix? Waere sehr gross!}
%%
All correlation values are scaled between 0 - no correlation and 1 - identical.
%%
Some variables are too sparse for calculating correlations, for example \emph{treatment of diabetes}, or \emph{medication against high blood pressure}.
%%
We display the resulting \emph{contingency matrix} using a heat map, mapping correlation values to color brightness with white for 0 and dark blue for 1 \cite{Klemm2014VIS}.
%%
We calculated the contingency matrix for all size groups and looked for correlations between image- and non-image derived variables.
%%
The resulting contingency matrices are available as supplementary material and show no strong correlation with any of the parameter.
%%
Only weak correlations could be found for \emph{Mean Curvature} with \emph{gender} (0.42), \emph{body size} (0.39) and \emph{number of born children} (0.29).
%%
One surprising result was the small correlation with \emph{torsion}, which correlated with almost no variables (p values between 0 and 0.05) and \emph{parkinson} (0.24).
%%
This observation brought us to the conclusion to incorporate more sophisticated data mining techniques to assess the influence of the image-derived parameters.

@@

\section{\uppercase{Evaluation of Decision Trees}}
\label{sec:DecisionTrees}
%%
\noindent As described before, correlation coefficients fail to infer back pain status based on lumbar spine canal curvature and torsion.
%%
We rely on predictive classification trained to obtain a complex rule set on how combination of the image-parameter explain non-image variables.
%%
Decision trees are a popular classification method in data mining for creating predictive models.
%%
Leaves of a decision tree represent class labels, branches represent feature conjunctions leading to the class labels.
%%
Decision trees are easy to understand and to read.
%%
They work with numerical as well as categorical data.
%%
This allows epidemiological domain experts to interpret the results without having deep knowledge about the algorithm creating the tree.
%%
Readability is only granted for small trees, complex structures with many branches are not desirable.
%%
Too many branches also impose to the data.
\com{Quelle Decision Tree}

@@

\subsection{Interactive Display of Decision Trees}
We have to create a decision tree for every non-image variable to see which one can be explained by image-derived variables.
%%
Since we have 134 non-image variables, the calculation yields a corresponding number of trees.
%%
Dividing the subjects into four groups using body-size, the number increases to 402 trees.
%%
To make the results cognitive feasible, we have to abstract the results of the classification.
%%------------
\subsubsection{Visualization of Classification Results}
\label{subsec:VisualizationOfClassificationResults}
%%
We follow the visual information-seeking mantra and want to provide an overview first, then details-on-demand \ref{shneiderman1996}.
%%
The optimal classification uses a few rules to precisely characterize the target variable.
%%
Therefore, we are interested in \emph{small trees} with a \emph{low classification error rate}.
%%
The two measures form the axis for a \emph{scatter plot} of the classification results.

@@

Many follow-up variables are sparse, such as medication of diabetes or reason of early retirement.
%%
The classification algorithm may produce higher accuracy for variables with less subjects due to the small sample size.
%%
This makes these results less reliable.
%%
We removed variables with less than 100 subjects, since they have to few subjects to be statistically significant.
%%
Furthermore, we map the number of subjects associated with a variable to point diameter in the scatter plot.
%%
This allows to instantly assess the reliability of the result.
%%

We apply a square root scale for the tree size axis to highlight data points with few decision rules.
%%
Outlier results with very large decision trees distort otherwise the resulting plot.
%%

@@

Dummy variables abstract a categorical variable with multiple manifestations into several binary variables.

@@

We however still want to be able to look display \emph{details-on-demand} \ref{shneiderman1996} and examine a decision tree in detail.

@@

This is realized by clicking on a entry on the visualization, which then displays the corresponding decision tree in detail (Fig.~\ref{fig:decision-tree} b).

>>>>

\section{\uppercase{Epidemiological Background}}
\label{sec:EpidemiologicalBackground}
%\noindent \com{Background. Epidemiological Workflow, focus on statistical resilience, image data hard to analyze due to the large amount of subjects, poor image quality and lack of methods}.
\noindent Epidemiology is the study of dissemination, causes and results of health-related states and events.
%%
Epidemiological reasoning relies on a strict statistically driven workflow \cite{Fletcher}:
%%
\begin{itemize}
%%
	\item Physicians formulate hypotheses based on observations made in their clinical practice.
%%
	\item To assess a hypothesis, epidemiologists compile a list of variables depicting it.
%%
	\item Statistical methods, such as regression analysis, assess the association of selected variables with the investigated disease.
%%
\end{itemize}
%%
Mutually dependent variables make this analysis challenging.
%%
Many diseases, such as different cancer types, are more likely with increasing age.
%%
When for example analyzing influences of nutrition to prostate cancer, the results need to be age-normalized.
%%
Age acts as an \emph{confounder} for prostate cancer.
%%
These \emph{confounding} variables are often hard to find.
%%
Statistical correlation does not imply causation--epidemiologists need to assess the medical soundness of the statistical results.
%%------------------------------
\subsection{Epidemiological Data}
Epidemiological data originates from a wide range of studies.
%%
The study type depends on the condition of interest.
%%
Most common are case-control studies, analyzing one specific disease and its influences.
%%
We focus on data from large scale cohort studies.
%%
These studies aim to collect as much data as possible for each subject.
%%
As a result, these data can be analyzed regarding many diseases and conditions.
%%

Epidemiological data is heterogenous and incomplete.
%%
For example women-specific questions or data about a disease treatment only for subjects suffering from this condition.
%%
Statistical analysis has to take missing data into account.
%%

@@

Modern cohort studies often comprise medical image data.
%%
These data are hard to analyze, since segmentation algorithms are not generally available and need to be custom-made for each body structure.
%%
Segmentation data is usually analyzed by abstracting it into key figures, such as diameters or distances.
%%
These numeric values can be compared with non-image variables to retrieve correlations.
%%--------------------------------
\subsection{Lower Back Pain}
%%
The lower (\emph{lumbar}) spine is the most stressed spine part.
%%
Lower back pain is one of the most frequent diseases in the western civilization.
%%
Epidemiologists assume associations between lumbar back pain and lifestyle factors.
%%
These include nutrition, sporting activities and body posture combined with physical stress at work.
%%
The exact causes as well as particularly vulnerable risk groups are not known.
%%
Potential \emph{confounding} effects are also subject of current research.
%%

Epidemiologists want to characterize the healthy aging process of the spine.
To achieve this, they have to analyze the lumbar spine shape as well as the mentioned lifestyle factors.
%% -------------------------------
\section{\uppercase{Related Work}}
\label{sec:RelatedWork}
\paragraph{Visual Analysis of Image and non-Image Data}
Klemm et al. \cite{Klemm2014VIS} propose an extension to the previously described epidemiological workflow using \emph{Interactive Visual Analysis} (IVA).
The workflow consists of an iterative sequence of \emph{group selection}, \emph{variable selection} and \emph{visualization}.
It aims to trigger \emph{hypothesis generation} by providing visualizations able to concurrently analyzing multiple heterogenous variables at once using correlation measures.
We incorporate the IVA approach by concurrently displaying complex relationships between variables using decision trees.
Steenwijk et al. \cite{Steenwijk} propose a coordinated linked view system for both image-related and non-image data.
It incorporates multiple plot types, such as scatter plots, parallel coordinates and time plots with brushing and linking facilities.
Similar to our work, they quantify image data project it into the non-image data information space.
Turkay et al. \cite{Turkay} follow a similar approach by deriving descriptive data metrics from image data.
Their proposed \emph{deviation plot} shows distribution-specific measures of a variable, such as skewness or inter-quantile-range, making variables comparable.
This approach aims to trigger \emph{hypothesis generation} by outlining tendencies between these variables.
\paragraph{Visual Analysis of Heterogenous Non-Image Data.}
Zhang et al. \cite{Zhang} analyze subject groups in a web-based linked view system.
The resulting decision rules aim to categorize new subjects as they are added to the data.
They define a cohort as parameter-divided subject group, differing from the epidemiological understanding of the term.
The paper lacks details about handling of missing data, the definition of similarity or which statistical measures are applied.
Generalized Pairs Plots (\texttt{GPLOM'S}) visualize heterogenous variables in a plot matrix \cite{GPLOMS,Francois}.
The plot depends on the type of variables, which are pairwise visualized.
The plot matrix is well suited for a overview visualization, but takes up much screen space and is therefore only suitable for few variables at once (see Fig.~\ref{fig:image-parameter-range}).
Dai et al. \cite{Dai} defined a \emph{Concept Map}, linking cancer-associated features by incorporating choropleth maps with bar charts, parallel coordinates and scatter plots with regression lines.
The \emph{Concept Map} is iteratively refined using primarily the underlying geographical data.
%Our work, however, does not contain geographical information of any kind.
%%----------------------------
\paragraph{Decision Rule Driven Analysis of Medical Data.}
Related work in this field often focuses on clinical diagnosis.
Closest to ours is the work of Gla{\ss}er \cite{Glasser2013} and Niemann et al. \cite{Niemann2014}.
Gla{\ss}er et al. \cite{Glasser2013} use variables derived from DCE-MRI image capturing the blood flow in the tumorous tissue.
They aim to classify breast tumors.
The extracted variables train a decision-tree classifier, concluding that the extracted features alone are not sufficient for tumor type classification.
Niemann et al. \cite{Niemann2014} assess hepatic steatosis (fatty liver disease) risk factors using decision trees.
They presented an interactive data mining tool, which can analyze association rules and highlight interesting relations, which may trigger new hypothesis about the data.
We combine ideas of both papers by validating the significance of image derived parameter aiming to derive new relationships by training decision trees to explain target variables.
%%----------------------------

Pinhero et al. \cite{Pinheiro} applied association rule mining to create high-confidence decision rules about the mortality rate in liver cancer data.
%%
Decision rules are easily readable and are well suited for classifying medical data. \com{Das frueher erwaehnen}
%%
The results can be easily communicated to domain experts with little knowledge about the algorithm behind the results.
%%
Once these rules, get too complex, analyzing these decision rules becomes complex.
%%
Sekhavat and Hoeber \cite{Sekhavat} propose a interactive visualization technique for decision trees, where all decisions are displayed in a heat map.
%%
The visualization translates the decision tree into a heat map, where each cell represents the connectivity between a initial value and an target value.
%%
The resulting interactive visualization is well suited for analyzing single, but not multiple decision trees.
%%
\paragraph{SHIP Data Analysis.}
Klemm et al. analyzed the lumbar spine variability of of 490 \texttt{SHIP-2} subjects \cite{Klemm2013VMV}.
%%
They incorporate hierarchical agglomerative clustering to derive shape groups, yielding several groups of average shape and several outlier cluster.
%%
The extracted lumbar spine shape was correlated with subject size.
%%
Out data extraction process is based on the detection algorithm and lumbar spine canal extraction presented in this paper.
%%
Clustering techniques applied to non-image data are strongly dependent on the chosen variables and distance measures.
%%
Klemm et al. found \texttt{k-Prototypes} and \texttt{DBSCAN} appropriate for this task \cite{Klemm2014BVM}.
%%
\\\\
Unique for our work is the combination of an IVA-approach by observing interesting feature relations in context of image-derived features using multiple decision trees for one visualization.
%%
We abstract the decision tree results similar to \cite{Turkay}, making them comparable in a overview visualization.
%%
\section{The Lumbar Spine Data Set}
\label{sec:MaterialsAndMethod}
%%
\noindent Our approach allows to analyze many variables simultaneously.
%%
Therefore, epidemiologists compiled the data set with a with a wide range of variables possibly correlating with lumbar back pain.
%%
The data set comprises of 6,753 subjects from two cohorts (4,420 from \texttt{SHIP-Trend-0} and 2,333 from \texttt{SHIP-2}).
%%
%It is an active research area to determine these correlations.
%%

%%
\subsection{Non-Image Data}
%\com{Analysis Using Heterogeneous Correlation}
The variables range from somatometric variables describing body measures to medical examinations, such as laboratory tests as well as lifestyle factors as sport activity or nutrition. \com{Ein paar Details hierzu noch - Datentyp, missing data, aufeinander aufbauende parameter}

@@

A hierarchical finite element method was used to detect the lumbar spine in the MRI data \cite{Rak2013}.
%%
The tetrahedron-based finite element model is initialized using three user-placed land marks.
%%
A click on the L3 vertebra center initialized the model, two clicks on the top and bottom of the vertebra describes an initial height estimation.
%%
The registered model captures information about the lumbar spine canal shape and the position of the L1-L5 vertebrae \cite{Klemm2013VMV} (Fig.~\ref{fig:centerline}~a).
%%
The detection fails for several subjects due imprecise initialization, imaging artifacts or strongly deformed spines.
%%
$2,540$ lumbar spine models were obtained and constitute the foundation of this work.
%%
%%
% \com{From VIS'14 Paper}
% The lumbar spine was detected in the image data using a hierarchical finite element method by Rak et al. [32].
% %%
% This semiautomatic method requires the user to initialize the tetrahedron-based finite element models (FEM) with a click on the L3 vertebra.
% %%
% Two user-defined landmarks on the top and bottom of the L3 vertebra describe an initial model height estimation.
% %%
% The model uses a weighted sum of T1 and T2-weighted MR images to detect the lumbar spine shape.
% %%
% Once registered, it captures information about the shape of the lumbar spine canal as well as the position of the L1-L5 vertebrae [21].
% %%
% Due to incorrect initialization, strongly deformed spines, contrast differences and artifacts, the model was not able to detect lumbar spines for all subjects.
% %%
% We obtained and worked with 2,540 tetrahedron models of the lumbar spine.
% %%
% For clustering, we extracted the centerline of the lumbar spine canal, which captures information about lordosis and scoliosis (the medical terms for spine curvature) [21].
% %%

We have to assess the model accuracy to extract key figures from it.
%%
The detection model depicts the vertebrae positions, but lacks detailed information about their volume.
%%
It captures reliable information about spine canal curvature.
%%
In a previous work \cite{Klemm2013VMV,Klemm2014VIS}, we extracted a centerline representation of the lumbar spine canal from the detection model (Fig.~\ref{fig:centerline}~a).
%%
Using the Frenet formulas \cite{Frenet}, we calculated the following metrics from the model (Fig.~\ref{fig:centerline}~b):
%%
%Thus, we extracted the following metrics from the model:
%%
%\paragraph{Curvature}
\begin{itemize}
	% \item The curvature of the centerline is calculated using the summed mean curvature between adjacent points:
% %%
% $\sum \frac{curvature_i}{curvature_all}$.
	\item \emph{Curvature} is calculated as weighted sum of curvature between all adjacent points describing the centerline: $\sum \frac{curvature_i}{curvature_{all}}$.
%%
	\item \emph{Torsion} (deviation of a curve from its current course) is calculated as weighted sum of torsion between all adjacent points describing the centerline: $\sum \frac{torsion_i}{torsion_{all}}$.
	\item \emph{Curvature angle $\alpha$} is defined by the middle point of the spine canal centerline as \emph{vertex} and the line between middle point and top/bottom point as \emph{sides}.
%%
\end{itemize}
%%
These figures are also extracted in the sagittal, coronal and transversal projection of the model.
%%
We assess the information gain of each dimension using these projections.
%%
This gives us a total of 9 image-derived parameters.

@@

Larger people have also a longer spine and its shape is more straight.

@@

\noindent As first experiment we correlated the shape parameter with the binary back pain indicator using \texttt{GPLOM's} \cite{GPLOMS}.

@@

%Pairs plots visualize continuous and categorical variables pairwise in a matrix view.

@@

%Scatterplots display pairs of continuous variables, bar charts for continuous variable as a function of a categorical variable and box plots pairs of categorical variables.

@@

\emph{Curvature}, \emph{curvature angle} and \emph{torsion} correlate strongly with their planar projections.

@@

Also the mean curvature and the curvature angle correlate by a factor of $-0.89$.

@@

\subsubsection{Assessing the Information Gain Using the PCA}
%%
To determine the information gain per image-derived variables, we calculated a PCA and compared the loadings per dimension.
%%
The first three principal components explain 75\% of variance in the image-derived variables.
%%
The first principal component explains 47\% of the variance and weights primarily \emph{mean curvature}, \emph{curvature sagittal}, \emph{curvature angle} and \emph{curvature angle sagittal}.
%%
The second component, adding 16\% of variance, weigths  \emph{curvature coronal} and \emph{curvature angle coronal}.
%%
The third component explains 12\% of variance and weights \emph{torsion} and \emph{curvature transverse}.
%%
This supports our prior conclusion about the low information gain of the transverse planes.
%%
\emph{Torsion} also adds little variance to the information space.
%% ------------------------------------------
\subsubsection{Heterogenous Correlations}
%%
We then expanded our focus on correlations of image-derived parameters with all other non-image variables.
%%
We applied a heterogenous correlation technique to derive correlations between all variables is the data set.
%%
The method uses the following correlation metrics for the different type combination:
%%
\begin{itemize}
%%
\item \emph{Pearson product-moment} for two continuous variables, 
\item \emph{Polyserial} correlation for one continuous and one categorical variable, and
\item \emph{Polychoric} correlations for two categorical variables.
\end{itemize}
\com{TODO: Abbildung einer Korrelationsmatrix? Waere sehr gross!}
%%
All correlation values are scaled between \texttt{0 - no correlation} and \texttt{1 - identical}.
%%
Some variables are too sparse for calculating correlations, for example \emph{treatment of diabetes}, or \emph{medication against high blood pressure}.
%%
We display the resulting \emph{contingency matrix} using a heat map, mapping correlation values to color brightness with white for 0 and dark blue for 1 \cite{Klemm2014VIS}.
%%
We calculated the contingency matrix for all size groups and looked for correlations between image- and non-image derived variables.
%%
The resulting contingency matrices are available as supplementary material and show no strong correlation with any of the parameter.
%%
Only weak correlations could be found for \emph{Mean Curvature} with \emph{gender} (0.42), \emph{body size} (0.39) and \emph{number of born children} (0.29).
%%
One surprising result was the small correlation with \emph{torsion}, which correlated with almost no variables (p values between 0 and 0.05) and \emph{parkinson} (0.24).
%%

These observations brought us to the conclusion to incorporate more sophisticated data mining techniques to assess the influence of the image-derived parameters.

@@

\section{\uppercase{Evaluation of Decision Trees}}
\label{sec:DecisionTrees}
%%
\noindent As described before, correlation coefficients fail to infer back pain status based on lumbar spine canal curvature and torsion.
%%
We rely on predictive classification trained to obtain a complex rule set on how combination of the image-parameter explain non-image variables.
%%
Decision trees are a popular classification method in data mining for creating predictive models.
%%
Leafs of a decision tree represent class labels, branches represent feature conjunctions leading to the class labels.
%%
Decision trees are easy to understand and to read.
%%
They work with numerical as well as categorical data.
%%
This allows epidemiological domain experts to interpret the results without having deep knowledge about the algorithm creating the tree.
%%
Readability is only granted for small trees, complex structures with many branches are not desirable.
%%
Too many branches also impose overfitting to the data.
\com{Quelle Decision Tree}

@@

\subsection{Interactive Display of Decision Trees}
We have to create a decision tree for every non-image variable to see which one can be explained by image-derived variables.
%%
Since we have 134 non-image variables, the calculation yields a corresponding number of trees.
%%
Dividing the subjects into four groups using body-size, the number increases to 402 trees.
%%
To make the results cognitive feasible, we have to abstract the results of the classification.
%%------------
\subsubsection{Visualization of Classification Results}
\label{subsec:VisualizationOfClassificationResults}
%%
We follow the visual information-seeking mantra and want to provide an overview first, then details-on-demand \cite{shneiderman1996}.
%%
The optimal classification uses a few rules to precisely characterize the target variable.
%%
Therefore, we are interested in \emph{small trees} with a \emph{low classification error rate}.
%%
The two measures form the axis for a \emph{scatter plot} of the classification results.

@@

Many follow-up variables are sparse, such as \emph{medication of diabetes} or \emph{reason of early retirement}.
%%
The classification algorithm may produce higher accuracy for variables with less subjects due to the small sample size.
%%
This makes these results less reliable.
%%
We removed variables with less than 100 subjects, since they have to few subjects to be statistically significant.
%%
Furthermore, we map the number of subjects associated with a variable to point diameter in the scatter plot.
%%
This allows to instantly assess the reliability of the result.
%%

We apply a square root scale for the tree size axis to highlight data points with few decision rules.
%%
Outlier results with very large decision trees would otherwise distort the resulting plot.
%%

@@

Dummy variables convert a categorical variable with multiple manifestations into several binary variables.

@@

We however still want to be able to look display \emph{details-on-demand} \cite{shneiderman1996} and examine a decision tree in detail.

@@

This is realized by clicking on a entry on the visualization, which then displays the corresponding decision tree in detail.
